Ground state energy of quantum dots using the coupled cluster method

### 2018-06-07 project_2/article/article.tex
Fix results for N = 12
Fix the last table
Add TODO's
Wrote more.

        The we can define a new function consisting of all of \autoref{eq:CCD_amp_1}
        except term number two and three,
        \begin{equation}
        \begin{aligned}
            g(u, \tau) =&
                    u^{ab}_{ij} + f^b_c t^{ac}_{ij}P(ab) - f^k_jt^{ab}_{ik}P(ij) \\
                 +& \frac{1}{4}t^{cd}_{ij} t^{ab}_{mn} u^{mn}_{cd} + \frac{1}{2}t^{cd}_{ij} u^{ab}_{cd} \\
                 +& \frac{1}{2}t^{cd}_{jm} t^{ab}_{in} u^{mn}_{cd} P(ij) - \frac{1}{2}t^{ac}_{nm} t^{bd}_{ij} u^{nm}_{cd} P(ab) \\
                 +& t^{ac}_{im} t^{bd}_{jn} u^{mn}_{cd} P(ij) + t^{ac}_{im} u^{bm}_{jc} P(ab) P(ij) \\
                 -& \frac{1}{2}t^{ab}_{im} u^{mn}_{jn}.
        \end{aligned}
        \end{equation}

@@

        Now we have $D_{ij}^{ab}t_{ij}^{ab} = g(u, \tau)$.
        This allows us to define an iterative scheme,
        \begin{equation}
            \tau^{(k + 1)} = \frac{g(u, \tau^{(k)})}{D_{ij}^{ab}},
            \label{eq:iterative_amplitude}
        \end{equation}
        with the initial guess
        \begin{equation}
            \tau^{(0)} = \frac{u_{ij}^{ab}}{D_{ij}^{ab}}.
        \end{equation}

        \subsubsection{Intermediate Computations}
        Looking closely at the amplitude equation in (\ref{eq:CCD_amp_1}) one might come
        to realize that it contains that this equation contains many of the same structures in
        several of the terms. This warrants the search for an algebraic transformation of the
        CCD amplitude equation that has the potential to reduce the amount of floating point
        operations needed to compute it. As it turns out, such terms exist and they will decrease
        the computing time necessary by an order of magnitude. We will define the following
        "intermediates",
        \begin{align}
            \chi^{ab}_{cd} &= \frac{1}{4}t^{ab}_{mn} u^{mn}_{cd} + \frac{1}{2}u^{ab}_{cd} \\
            \chi^n_j &= \frac{1}{2}t^{cd}_{jm} u^{mn}_{cd} \\
            \chi^a_d &= \frac{1}{2} t^{ac}_{nm} u^{nm}_{cd} \\
            \chi^{bm}_{jc} &= u^{bm}_{jc} + \frac{1}{2}t^{bd}_{jn}u^{mn}_{cd}
        \end{align}
        These intermediate structures will allow us to rewrite \autoref{eq:CCD_amp_1} to,
        \begin{equation}
        \begin{aligned}
            0 =& u^{ab}_{ij} + f^b_c t^{ac}_{ij}P(ab) - f^k_jt^{ab}_{ik}P(ij) \\
              &+ t^{cd}_{ij}\chi^{ab}_{cd} + t^{ab}_{in}\chi^n_jP(ij)
               - t^{bd}_{ij}\chi^a_dP(ab) \\
              &+ t^{ac}_{im}\chi^{bm}_{jc}P(ab)P(ij) + \frac{1}{2}t^{ab}_{im}u^{mn}_{jn}.
        \end{aligned}
        \end{equation}
        The importance of this "trick" will become apparent in due time.

@@

            Introducing a basis set transforms the Hartree-Fock equations into the Rothaan
            equations
            \begin{equation}
                FC = SC\varepsilon.
            \end{equation}
            This is a generalised eigenvalue problem where $S$ serves as an overlap matrix
            that must be there in case of non-orthogonal basis. Since the Fock matrix $F$
            depends on it's own solution through the orbitals, the eigenvalue problem must
            be solved iteratively\footnote{This is also the reason why the
            Hartree-Fock-Roothaan equations are often called the self-consistent-field
            procedure.}.

\section{Implementation}
    We have costructed a flexible framework for making performing CCD
    computations, consisting of package for \lstinline{python} that is easy
    to install globally on any computer.

    \subsection{Installation}
        The software is easy to install by first cloning the GitHub reopsitory,
        \begin{bash}
            git clone git@github.com:Schoyen/FYS4411.git
        \end{bash}
        Then all you need to do is change directory to the project code folder
        where a \lstinline{MakeFile} is included so you only need to build and
        install,
        \begin{bash}
            cd FYS4411/project_2/coupled-cluster
            make build
            make install
        \end{bash}
        Now everything should be able to run from anywhere on your computer.
        To ensure that all requirements of the program are satified you can
        install with \lstinline{make installr} instead.

@@

        %% NONE OF THIS WORKED AS INTENDED.
        %\begin{figure}
        %\centering
        %\framebox[0.5\textwidth]{%
        %\begin{minipage}{0.49\textwidth}
        %\begin{center}
        %\begin{varwidth}{0.49\textwidth}
        %\dirtree{%
        %.1 coupled\_cluster.
        %    .2 matrix\_elements.
        %        .3 generate\_matrices.
        %        .3 index\_map.
        %    .2 hartree\_fock.
        %        .3 scf\_rhf.
        %        .3 basis\_transformation.
        %    .2 schemes.
        %        .3 ccd.
        %        .3 ccd\_sparse.
        %        .3 ccd\_optimized.
        %}
        %\end{varwidth}
        %\end{center}
        %\end{minipage}
        %}
        %\end{figure}

@@

    \subsection{Mixing}
        A way to try ``massage'' convergence out of the CCD-method is to use
        \emph{mixing}. A popular solution is Pulay mixing\footnote{Commonly
        known as DIIS (direct inversion in the iterative subspace).}. We did not
        implement Pulay mixing but used a more naïve and simpler solution.
        \begin{align}
            \tilde{t}^{k + 1} = \theta t^{k + 1} + (1 - \theta)t^k,
        \end{align}
        where $t^{k + 1}$ is the current value computed using
        \autoref{eq:iterative_amplitude} and $t^k$ is the previous value for the
        amplitude. Choosing $\theta \in [0, 1]$ we can tune how much of the
        previous amplitude we wish to include in the new state. This allows for
        a more gradual transition between the iterations. We now use
        $\tilde{t}^{k + 1}$ as our estimate of the new amplitude.

@@

    \begin{table}
        \centering
        \caption{N = 2}
        \begin{ruledtabular}
            \begin{tabular}{c|c|ccc}
                $\omega$ & $R$ & RHF & CCD(HO) & CCD(HF) \\
                \hline
                           &  $1$  & $0.596333$ & $0.596333$ & $0.596333$ \\
                           &  $2$  & $0.596333$ & $0.512520$ & $0.512520$ \\
                           &  $3$  & $0.526903$ & $0.505972$ & $0.442235$ \\
                           &  $4$  & $0.526903$ & $0.499216$ & $0.442011$ \\
                           &  $5$  & $0.525666$ & $0.497172$ & $0.443293$ \\
    \multirow{2}{*}{$0.1$} &  $6$  & $0.525666$ & $0.494232$ & $0.443145$ \\
                           &  $7$  & $0.525635$ & $0.493142$ & $0.443056$ \\
                           &  $8$  & $0.525635$ & $0.491895$ & $0.442981$ \\
                           &  $9$  & $0.525635$ & $0.491262$ & $0.442927$ \\
                           &  $10$ & $0.525635$ & $0.490649$ & $0.442886$ \\
                           &  $11$ & $0.525635$ & $0.490262$ & $0.442853$ \\
                           &  $12$ & $0.525635$ & $0.489918$ & $0.442827$ \\
                \hline

@@

    \begin{table}
        \centering
        \caption{N = 6}
        \begin{ruledtabular}
            \begin{tabular}{c|c|ccc}
                $\omega$ & $R$ & RHF & CCD(HO) & CCD(HF) \\ \hline
                         &  $2$  & $4.864244$ & $4.864244$ & $4.864244$ \\
                         &  $3$  & $4.435740$ & $4.446235$ & $4.319901$ \\
                         &  $4$  & $4.019787$ & $4.383692$ & $3.829962$ \\
                         &  $5$  & $3.963149$ & \nan & $3.666722$ \\
                         &  $6$  & $3.870617$ & \nan & $3.597876$ \\
                  $0.1$  &  $7$  & $3.863135$ & \nan & $3.590388$ \\
                         &  $8$  & $3.852880$ & \nan & $3.587711$ \\
                         &  $9$  & $3.852591$ & \nan & $3.587291$ \\
                         &  $10$ & $3.852393$ & \nan & $3.587136$ \\
                         &  $11$ & $3.852391$ & \nan & $3.586821$ \\
                         &  $12$ & $3.852382$ & \nan & $3.586582$ \\ \hline

@@

    \subsection{Validity of results}
        All of our results have been compared with the master thesis of M. P.
        Lohne\cite{lohne2010coupled} for up to 10 shells. Lohne gets two sets of
        energies, one set from RHF and one set from CCSD code with harmonic
        oscillator basis functions. Our CCD code with Hartree-Fock basis will
        for some configurations\footnote{By configurations we mean frequency
        $\omega$ and number of particles $N$.} beat CCSD with plain harmonic
        oscillator basis. For 12 shells we can compare with the results from M.
        P. Lohne et al.\cite{lohne2011ab}, but in this article an
        \emph{effective interaction} for the Hamiltonian with a CCSD code has
        been used. These results are therefore significanlty better than ours,
        but we can get a ``ball-park'' idea to benchmark against.

@@

        We can see this behaviour in our results. For $N = 2$ we get convergence
        for all methods for $\omega = 0.1$ shown in \autoref{tab:N2}. By
        increasing to $N = 6$ particles we immediately see how CCD with the HO
        basis will get in trouble for small values of $\omega$. By increasing
        the strength of the confinment even this method does converge.

@@

@@

>>>>

        The we can define a new function consisting of all of \autoref{eq:CCD_amp_1}
        except term number two and three,
        \begin{equation}
        \begin{aligned}
            g(u, t) =&
                    u^{ab}_{ij} + f^b_c t^{ac}_{ij}P(ab) - f^k_jt^{ab}_{ik}P(ij) \\
                 +& \frac{1}{4}t^{cd}_{ij} t^{ab}_{mn} u^{mn}_{cd} + \frac{1}{2}t^{cd}_{ij} u^{ab}_{cd} \\
                 +& \frac{1}{2}t^{cd}_{jm} t^{ab}_{in} u^{mn}_{cd} P(ij) - \frac{1}{2}t^{ac}_{nm} t^{bd}_{ij} u^{nm}_{cd} P(ab) \\
                 +& t^{ac}_{im} t^{bd}_{jn} u^{mn}_{cd} P(ij) + t^{ac}_{im} u^{bm}_{jc} P(ab) P(ij) \\
                 -& \frac{1}{2}t^{ab}_{im} u^{mn}_{jn}.
        \end{aligned}
        \end{equation}

@@

        Now we have $D_{ij}^{ab}t_{ij}^{ab} = g(u, t)$.
        This allows us to define an iterative scheme,
        \begin{equation}
            t^{(k + 1)} = \frac{g(u, t^{(k)})}{D_{ij}^{ab}},
            \label{eq:iterative_amplitude}
        \end{equation}
        with the initial guess
        \begin{equation}
            t^{(0)} = \frac{u_{ij}^{ab}}{D_{ij}^{ab}}.
        \end{equation}

        \subsubsection{Intermediate Computations}
        Looking closely at the amplitude equation in (\ref{eq:CCD_amp_1}) one might come
        to realize that it contains that this equation contains many of the same structures in
        several of the terms. This warrants the search for an algebraic transformation of the
        CCD amplitude equation that has the potential to reduce the amount of floating point
        operations needed to compute it. As it turns out, such terms exist and they will decrease
        the computing time necessary by an order of magnitude. We will define the following
        "intermediates",
        \begin{align}
            \label{eq:intermediate1}
            \chi^{ab}_{cd} &= \frac{1}{4}t^{ab}_{mn} u^{mn}_{cd} + \frac{1}{2}u^{ab}_{cd} \\
            \label{eq:intermediate2}
            \chi^n_j &= \frac{1}{2}t^{cd}_{jm} u^{mn}_{cd} \\
            \label{eq:intermediate3}
            \chi^a_d &= \frac{1}{2} t^{ac}_{nm} u^{nm}_{cd} \\
            \label{eq:intermediate4}
            \chi^{bm}_{jc} &= u^{bm}_{jc} + \frac{1}{2}t^{bd}_{jn}u^{mn}_{cd}
        \end{align}
        These intermediate structures will allow us to rewrite \autoref{eq:CCD_amp_1} to,
        \begin{equation}
        \begin{aligned}
            0 =& u^{ab}_{ij} + f^b_c t^{ac}_{ij}P(ab) - f^k_jt^{ab}_{ik}P(ij) \\
              &+ t^{cd}_{ij}\chi^{ab}_{cd} + t^{ab}_{in}\chi^n_jP(ij)
               - t^{bd}_{ij}\chi^a_dP(ab) \\
              &+ t^{ac}_{im}\chi^{bm}_{jc}P(ab)P(ij) + \frac{1}{2}t^{ab}_{im}u^{mn}_{jn}.
        \end{aligned}
        \end{equation}
        The importance of this "trick" will become apparent in due time.

@@

            Introducing a basis set transforms the Hartree-Fock equations into the Rothaan
            equations
            \begin{equation}
                FC = SC\varepsilon.
            \end{equation}
            This is a generalised eigenvalue problem where $S$ serves as an overlap matrix
            that must be there in case of non-orthogonal basis. Since the Fock matrix $F$
            depends on it's own solution through the orbitals, the eigenvalue problem must
            be solved iteratively\footnote{This is also the reason why the
            Hartree-Fock-Roothaan equations are often called the self-consistent-field
            procedure.}

\section{Implementation}
    We have costructed a flexible framework for making performing Hartree-Fock 
    self-consistednt field (SCF) and coupled cluster with double excitations (CCD)
    computations. The entire code base consists of package for \lstinline{python}
    that is easy to install globally on any computer.

    Because we have employed a mixture of Python and C++ with Cython interfaces,
    the code sturcture may be difficult to undertand for someone without Cython
    experience, but we have neverteless tried to make the usage of the software as
    painless as possible.

    \subsection{Installation and Usage}

        The software is easy to install by first cloning the GitHub reopsitory,
        \begin{bash}
            git clone git@github.com:Schoyen/FYS4411.git
        \end{bash}
        Then all you need to do is change directory to the project code folder
        where a \lstinline{MakeFile} is included so you only need to build and
        install,
        \begin{bash}
            cd FYS4411/project_2/coupled-cluster
            make build
            make install
        \end{bash}
        Now everything should be able to run from anywhere on your computer.
        To ensure that all requirements of the program are satified you can
        install with \lstinline{make installr} instead.

        We have included a sample python script in \autoref{app:sample_script}
        that uses almost all methods and classes in the package.

@@

@@

        Both in \mintinline{python}{CoupledClusterDoublesSparse} and
        \mintinline{python}{CoupledClusterDoublesOptimized} we make use
        of the intermediates in equations \ref{eq:intermediate1} to
        \ref{eq:intermediate4}. In the optimized class, special care was
        taken to compute

    \subsection{Convergence Problems and Mixing}

        Iterative many-body methods are prone to convergence problems for some
        configurations. Since Hartree-Fock is a variational method, SCF convergence
        is found when the energy is stationary with respect to inifitesimal 
        variations in the orbitals. Unfortunately, the SCF iteration scheme does 
        not always converge. Luckily, numerous techniques exist for controlling
        and accelerating convergence\cite{schlegel1991you}. The same kind of 
        methods have proven useful to ensure convergence of coupled cluster methods
        \cite{scuseria1986accelerating}.

        The simplest way to try to ``massage'' convergence out of the CCD-method 
        is to use \emph{damping} where you include a part of the result from the
        previous iteration i.e.,
        \begin{align}
            \tilde{t}^{k + 1} = \theta t^{k + 1} + (1 - \theta)t^k,
            \label{eq:mixing}
        \end{align}
        where $t^{k + 1}$ is the current value computed using
        \autoref{eq:iterative_amplitude} and $t^k$ is the previous value for the
        amplitude. Choosing $\theta \in [0, 1]$ we can tune how much of the
        previous amplitude we wish to include in the new state. This allows for
        a more gradual transition between the iterations. We now use
        $\tilde{t}^{k + 1}$ as our estimate of the new amplitude.

        A more sophisticated mixing method is the direct inversion 
        in the iterative subspace (DIIS), also known as Pulay mixing. While the
        common mixing method is used for many other applications
        \footnote{It is, for instance, called an alpha filter in data acquisition.},
        the DIIS method is developed with the sole intent of accelerating
        convergense in Hartree-Fock methods. In DIIS one would construct a
        linear combinations of approximate errors from previous iterations,
        analogous to a very clever weighted moving average. We have not implemented
        this scheme, but it nevertheless warrants mention.

@@

    \begin{table}
        \centering
        \caption{Results for N = 2 particles, where convergence was 
            acchieved for all parameters. Taut's\cite{taut1994two}
            analytic result for $\omega=1.0a.u.$ is $3$.}
        \begin{ruledtabular}
            \begin{tabular}{c|c|ccc}
                $\omega$ & $R$ & RHF & CCD(HO) & CCD(HF) \\
                \hline
                           &  $1$  & $0.596333$ & $0.596333$ & $0.596333$ \\
                           &  $2$  & $0.596333$ & $0.512520$ & $0.512520$ \\
                           &  $3$  & $0.526903$ & $0.505972$ & $0.442235$ \\
                           &  $4$  & $0.526903$ & $0.499216$ & $0.442011$ \\
                           &  $5$  & $0.525666$ & $0.497172$ & $0.443293$ \\
    \multirow{2}{*}{$0.1$} &  $6$  & $0.525666$ & $0.494232$ & $0.443145$ \\
                           &  $7$  & $0.525635$ & $0.493142$ & $0.443056$ \\
                           &  $8$  & $0.525635$ & $0.491895$ & $0.442981$ \\
                           &  $9$  & $0.525635$ & $0.491262$ & $0.442927$ \\
                           &  $10$ & $0.525635$ & $0.490649$ & $0.442886$ \\
                           &  $11$ & $0.525635$ & $0.490262$ & $0.442853$ \\
                           &  $12$ & $0.525635$ & $0.489918$ & $0.442827$ \\
                \hline

@@

    \begin{table}
        \centering
        \caption{Results for $N = 6$ particles. No convergence for HO
            basis for low frequency and large number of shells.}
        \begin{ruledtabular}
            \begin{tabular}{c|c|ccc}
                $\omega$ & $R$ & RHF & CCD(HO) & CCD(HF) \\ \hline
                         &  $2$  & $4.864244$ & $4.864244$ & $4.864244$ \\
                         &  $3$  & $4.435740$ & $4.446235$ & $4.319901$ \\
                         &  $4$  & $4.019787$ & $4.383692$ & $3.829962$ \\
                         &  $5$  & $3.963149$ & \nan & $3.666722$ \\
                         &  $6$  & $3.870617$ & \nan & $3.597876$ \\
                  $0.1$  &  $7$  & $3.863135$ & \nan & $3.590388$ \\
                         &  $8$  & $3.852880$ & \nan & $3.587711$ \\
                         &  $9$  & $3.852591$ & \nan & $3.587291$ \\
                         &  $10$ & $3.852393$ & \nan & $3.587136$ \\
                         &  $11$ & $3.852391$ & \nan & $3.586821$ \\
                         &  $12$ & $3.852382$ & \nan & $3.586582$ \\ \hline

@@

    \subsection{Validity of results}
        The analytical computed energy for $\omega=1.0a.u.$
        and $N=2$ is $3$\cite{taut1994two}. From our tables we see 
        that RHF is least successful in reaching this value, CCD in
        HF basis is best, while CCD with HO basis is generally somewhere 
        between the two. We see that we get even closer to the analytical
        ``truth'' by increasing the number of shells.

        The rest of results have been compared with the master thesis of M. P.
        Lohne\cite{lohne2010coupled} for up to 10 shells. Lohne gets two sets of
        energies, one set from RHF and one set from CCSD code with harmonic
        oscillator basis functions. Our CCD code with Hartree-Fock basis will
        for some configurations\footnote{By configurations we mean frequency
        $\omega$ and number of particles $N$.} beat CCSD with plain harmonic
        oscillator basis. For 12 shells we can compare with the results from M.
        P. Lohne et al.\cite{lohne2011ab}, but in this article an
        \emph{effective interaction} for the Hamiltonian with a CCSD code has
        been used. These results are therefore significanlty better than ours,
        but we can get a ``ball-park'' idea to benchmark against.

@@

        We can see this behaviour in our results. For $N = 2$ we get convergence
        for all methods for $\omega = 0.1$ shown in \autoref{tab:N2}. By
        increasing to $N = 6$ particles we immediately see how CCD with the HO
        basis will get in trouble for small values of $\omega$. Even by increasing
        the strength of the confinment this method does converge.

@@

        CCD with HO basis experiences the same convergence problems also for,
        $N=12$ particles and for $N=20$ particles.

@@

    \label{app:sample_script}
